Statistical Methods

### 2022-07-13 notes/theory.tex
cleaned up notes a bit

The total size of the set of samples $\mathcal{S}$ is $n_\mathcal{S} =
|\mathcal{S}|$. Assuming all sites are biallelic, we can simplify the
calculation of the likelihood by noting that for a site $v$'s vector of allele
counts $[c_1, c_2]$, the total number of pairwise combinations with the same
alleles is

@@

where $n_T = n_\mathcal{S} (n_\mathcal{S} - 1) / 2$  is the total number of
pairwise combinations across the sample set $\mathcal{S}$. Furthermore, note
that we can partition the set $\mathcal{V}$ of neutral sites into polymorphic
($\mathcal{P}$) and fixed sites ($\mathcal{F}$), i.e. $\mathcal{V} =
\mathcal{P} \cup \mathcal{F}$ and $\mathcal{P} \cap \mathcal{F} = \varnothing$.
For all $v \in \mathcal{F}$, $n_d(v) = 0$ and $n_s(v) = n_T$.

@@

In practice, we calculate these values across bins. For each bin, we treat
$\pi(v | \theta)$ as fixed, assuming that at this scale, the variation in
expected diversity across sites is minimal. For a particular chromosome, we
have two classes of sites: those included in the diversity calculation and
those ignored. The former sites are all putatively neutral and have reliably
called genotypes, and the other sites are possibly non-neutral or do have
reliably called genotypes. The log-likelihood is the sum of bin
likelihoods, $\mathcal{L}(b)$

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{P}} n_\text{D}(v) + \log(1-\bar{\pi}(b | \theta)) \left(\sum_{v \in \mathcal{V}} n_\text{S}(v) +  \sum_{v \in \mathcal{F}} n_T(v)  \right).
\end{align}

We can assume that the total number of combinations at each fixed site is
constant, e.g. $n_T = n_T(v)$ for all $v$, simplifying this as

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{P}} n_\text{D}(v) + \log(1-\bar{\pi}(b | \theta)) \left(\sum_{v \in \mathcal{V}} n_\text{S}(v) +  n_T(v) f(b)  \right)
\end{align}

where $f(b)$ is the number of fixed sites in bin $b$. Thus, we our data can be
reduced within a bin by these terms summed over sites in the window,

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) Y_D(b) + \log(1-\bar{\pi}(b | \theta)) Y_S(b)
\end{align}

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}_0 B(b | \vec{w}, \vec{t})) Y_D(b) + \log(1-\bar{\pi}_0 B(b |  \vec{w}, \vec{t})) Y_S(b)
\end{align}

where $\vec{w}$ and $\vec{t}$ are the vectors of mutation rates into

There are two approaches: interpolated $\pi(m | \theta)$ at the midpoint of the
bin position $m$, and averaged bin-averaged $\bar{\pi}(m | \theta)$.

>>>>

The total size of the set of samples $\mathcal{S}$ is $n_\mathcal{S} =
|\mathcal{S}|$. Assuming all sites are biallelic, we can simplify the inner
summation by counting the number of possible same and different pairwise
combinations. If a site $v$'s vector of allele counts is $[c_1, c_2]$, the
total number of pairwise combinations with the same alleles is

@@

where $n_T = n_\mathcal{S} (n_\mathcal{S} - 1) / 2$  is the total number of
pairwise combinations across the sample set $\mathcal{S}$. Note that these
site-specific counts allow us to use allelic counts directly, and can vary
across sites. Our likelihood is then,

\begin{align}
  \log\mathcal{L}(\theta) &= \sum_{v \in \mathcal{V}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right].
\end{align}

In practice, we calculate these values across bins. For each bin, we treat
$\pi(v | \theta)$ as fixed, assuming that at this scale, the variation in
expected diversity across sites is minimal. For a particular chromosome, we
have two classes of sites: those included in the diversity calculation and
those ignored. The former sites are all putatively neutral and have reliably
called genotypes, and the other sites are possibly non-neutral or do have
reliably called genotypes. The total log-likelihood is the sum of bin likelihoods,
$\mathcal{L}(b)$

\begin{align}
   \mathcal{L}(\theta) =  \sum_b \mathcal{L}(b | \theta)
\end{align}

The likelihood within a bin is then,

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{V}_b} n_D(v) + \log(1-\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{V}_b} n_S(v)  \\
                               &= \log(\bar{\pi}(b | \theta)) Y_D(b) + \log(1-\bar{\pi}(b | \theta)) Y_S(b)
\end{align}
%
where the two sum terms as $Y_D(b)$ and $Y_S(b)$ are data reductions at the bin
level.

If the data are such that only polymorphic sites are considered, we can adapt
this by partitioning the set $\mathcal{V}$ of neutral sites into polymorphic
($\mathcal{P}$) and fixed sites ($\mathcal{F}$), i.e. $\mathcal{V} =
\mathcal{P} \cup \mathcal{F}$ and $\mathcal{P} \cap \mathcal{F} = \varnothing$.
For all $v \in \mathcal{F}$, $n_d(v) = 0$ and $n_s(v) = n_T$.

@@

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{P}_b} n_\text{D}(v) + \log(1-\bar{\pi}(b | \theta)) \left(\sum_{v \in \mathcal{P}_b} n_\text{S}(v) +  \sum_{v \in \mathcal{F}_b} n_T(v)  \right).
\end{align}

Note that if we assume that the total number of combinations at each fixed site
is constant, e.g. $n_T = n_T(v)$ for all $v$, then we can use $\sum_v n_T(v) =
n_T |\mathcal{F}_b|$.

% There are two approaches: interpolated $\pi(m | \theta)$ at the midpoint of the
% bin position $m$, and averaged bin-averaged $\bar{\pi}(m | \theta)$.

\section*{Windowed Diversity}

Although we use the components of diversity, $n_t$ and $n_s$, to calculate the
likelihood, it is still of interest to calculate diversity from these in a
window. The raw allele count data a $L \times \text{2}$ matrix,
